conmat: generate synthetic contact matrices for a given age population

### 2024-05-30 paper/paper.Rmd
Added ORCID numbers

---
title: 'conmat: generate synthetic contact matrices for a given age population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 
date: "14 April 2023"
output:
  html_document:
    keep_md: yes
  pdf_document: default
bibliography: paper.bib
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
---

@@

Epidemiologists and public policy makers need to understand the spread of infectious diseases in a population. Knowing which groups are most vulnerable and how disease spread will unfold helps to make proactive decisions that benefit the most people.  Diseases like influenza and coronavirus spread via human-to-human, "social contact".  If we can measure the amount of social contact, we can use this to understand how diseases spread.

We can measure social contact through social contact surveys, where people  describe the number and type of social contact they have. These surveys provide the empirical estimates of the number of social contacts from one age group to another, as well as the setting of contact. For example, homes have higher contact with 25-50 year olds with 0-15 year olds, workplaces have high contact within 25-60 year olds.

These surveys exist for a variety of countries, for example, Mossong et al., the "POLYMOD" study, covered 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands and Poland. However, what do we do when we want to look at contact rates in different countries that haven’t been measured? We can use this existing data to help us project to countries or places that do not have empirical survey data. These are called "synthetic contact matrices". A very popular approach by Prem et al projected from the POLYMOD study to 177 countries, and provided contact matrices at "urban" and "rural" levels for each country.

However not all countries were included in this list from Prem et al, and for some analyses these synthetic contact matrices were at a large area, such as "urban" or "rural" for a given country. We needed to make predictions at finer grained regions, were required. It was challenging to apply the methodology used by Prem et al, as while their analysis code was provided, it was not designed for reuse. We needed to create contact matrices based on existing studies like POLYMOD but projected to the age population in a given local government area in Australia.

Conmat was created to fill a specific need for creating synthetic contact matrices for specific local government areas for Australia, for work commissioned by the Australian government. We created methods and software to facilitate the following:

@@

```{r}
perth <- abs_age_lga("Perth (C)")
perth
```

@@

```{r}
perth_contact <- extrapolate_polymod(population = perth)
perth_contact
```

@@

```{r}
autoplot(perth_contact)
```

@@

We provide multiple levels for the user to interact with for model fitting, further detail can be seen at: https://njtierney.github.io/conmat/dev/

@@

* Create a contact matrix using a custom contact survey from another source, such as the `socialmixr` R package.
* Predict to any age brackets - such as monthly ages, for example, 1, 3, 6, month year old infants
* Add ability to fit multiple contact surveys at once, e.g., POLYMOD and another dataset
* Add ability to include known household age distributions as offsets in the 'home' setting model, in place of the whole population distribution. So compute household age matrices (like age-structured contact matrices, but for household members instead of contacts) from Polymod data. If we compute a different one for each household size, in the Polymod data (probably estimated with another GAM, to make best use of the limited data) we might be able to extrapolate household age matrices to new countries based on the distribution of household sizes.
* Add methods for including household size distributions
* Add uncertainty to estimates
* Move Australian centric data into its own package
* Add documentation on specifying your own GAM model and using this workflow

@@

- Prem 2017: https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005697
- Prem 2021: https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1009098
- Mossonget al 2007: https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.0050074

>>>>

---
title: 'conmat: generate synthetic contact matrices for a given age population'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
- affiliation: 1,2
  name: Nick Golding
  orcid: 0000-0001-8916-5570
- affiliation: 1,3
  name: Aarathy Babu
  orcid: 
- affiliation: 4
  name: Michael Lydeamore
  orcid: 0000-0001-6515-827X
- affiliation: 1,3
  name: Chitra Saraswati
  orcid: 0000-0002-8159-0414
date: "03 May 2024"
output:
  html_document:
    keep_md: yes
  pdf_document: default
bibliography: references.bib
tags:
- epidemiology
- R
- infectious disease
affiliations:
- index: 1
  name: Telethon Kids Institute
- index: 2
  name: Curtin University
- index: 3
  name: 
- index: 4
  name: Monash University
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
#| comment: false
knitr::opts_chunk$set(comment = "#>",
                      echo = TRUE,
                      out.width = "95%",
                      retina = 3,
                      fig.height = 4,
                      fig.align = "center",
                      dpi = 300,
                      dev = "png")
options(tinytex.clean = FALSE)
```

@@

Epidemiologists and public policy makers need to understand the spread of infectious diseases in a population. Knowing which groups are most vulnerable, and how disease spread will unfold facilitates public health decision-making. Diseases like influenza and coronavirus spread via human-to-human, "social contact".  If we can measure the amount of social contact, we can use this to understand how diseases spread.

We can measure social contact through social contact surveys, where people  describe the number and type of social contact they have. These surveys provide an empirical estimate of the number of social contacts from one age group to another, as well as the setting of contact. For example, we might learn from a contact survey that homes have higher contact with 25-50 year olds and with 0-15 year olds, whereas workplaces might have high contact within 25-60 year olds.

These surveys exist for a variety of countries, for example, @mossong2008 the "POLYMOD" study, covered 8 European countries: Belgium, Germany, Finland, Great Britain, Italy, Luxembourg, The Netherlands and Poland [@mossong2008]. However, what do we do when we want to look at contact rates in different countries that haven’t been measured? We can use this existing data to help us project to countries or places that do not have empirical survey data. These are called "synthetic contact matrices". A very popular approach by Prem et al projected from the POLYMOD study to 152 countries [@prem2017]. This which was later updated to include contact matrices for 177 countries at "urban" and "rural" levels for each country [@prem2021].

However not all countries were included in this list from @prem2021, and for some analyses these synthetic contact matrices were at a large area, such as "urban" or "rural" for a given country. One limitation of this is that the predictions were not provided at smaller grained regons. This might be important, if for example, a public health group needed to make predictions for specific areas. It was challenging to apply the methodology used by Prem et al, as while their analysis code was provided, it was not designed for reuse.

The `conmat` package was created to fill a specific need for creating synthetic contact matrices for specific local government areas for Australia, for work commissioned by the Australian government. We created methods and software to facilitate the following:

@@

```{r}
#| label: load-conmat
library(conmat)
perth <- abs_age_lga("Perth (C)")
perth
```

@@

```{r}
#| label: extrapolate-polymod
#| echo: true
perth_contact <- extrapolate_polymod(population = perth)
perth_contact
```

@@

```{r}
#| label: autoplot-contacts
autoplot(perth_contact)
```

@@

We provide multiple levels for the user to interact with for model fitting, further detail can be seen at: https://idem-lab.github.io/conmat/dev/

@@

* Create a contact matrix using a custom contact survey from another source, such as the `socialmixr` R package.
* Predict to any age brackets - such as monthly ages, for example, 1, 3, 6, month year old infants
* Add ability to fit multiple contact surveys at once, e.g., POLYMOD and another dataset
* Add ability to include known household age distributions as offsets in the 'home' setting model, in place of the whole population distribution. So compute household age matrices (like age-structured contact matrices, but for household members instead of contacts) from POLYMOD data. If we compute a different one for each household size, in the POLYMOD data (probably estimated with another GAM, to make best use of the limited data) we might be able to extrapolate household age matrices to new countries based on the distribution of household sizes.
* Add methods for including household size distributions
* Add uncertainty to estimates
* Move Australian centric data into its own package
* Add documentation on specifying your own GAM model and using this workflow

@@
